---
title: "Remove Stimulation Artifacts"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
# This code block sets up the engine environment
# Please do not remove me
ravepipeline::pipeline_setup_rmd("stim_artifact_removal")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)
```

# Overview

Electrical stimulation delivered through an intracranial electrode injects a large,
stereotyped voltage transient into every recording channel. Each pulse produces two
things we want to remove:

1. a **direct artifact** — the electrode network saturates for a fraction of a
   millisecond around the pulse; and
2. a **secondary artifact** — the tissue/amplifier "rebounds" in the tens-of-
   milliseconds *after* each pulse, riding on top of the brain signal we care about.

This pipeline is a mathematically faithful R port of the MATLAB prototype
`tmp/artifact_rej.m`, wrapped in the standard RAVE "load many, analyze one" workflow.
The cleaning strategy has two stages, applied in order:

* **Bipolar re-reference** (within a shaft, skipping the stim contact) cancels the
  shared secondary artifact and common-mode noise for channels away from the stim site.
* **Template subtraction** removes the residual, near-site secondary artifact that
  survives bipolar re-referencing, by estimating the stereotyped per-pulse waveform and
  subtracting it pulse-by-pulse, then interpolating across the blanked direct-artifact
  region.

Pulse timing is recovered by thresholding the **stim electrode itself**: it carries the
cleanest square pulses and therefore the most reliable per-pulse timestamps. Because the
stim electrode can differ block-to-block, the pipeline analyzes **one block at a time**
(`analysis_block` + its `stim_electrode`); to process another block, change those two
settings and re-run.

The workflow is organized in four sections, mirroring the MATLAB script section-by-section.

# Section 1 — Load raw voltage

Artifact removal must operate on **raw** voltage: not referenced, not notch-filtered, not
epoched. The artifact lives in the raw trace, and any prior referencing would already have
mixed it across channels. We load a user-chosen subset of contacts (`loaded_electrodes`,
blank = all LFP channels) across the requested blocks (`loaded_blocks`, blank = all), using
`prepare_subject_raw_voltage_with_blocks`, which pages whole-block voltage from SSD via
`filearray`. The stim electrode is fetched on demand by `load_stim_channel()`, which reads
it from the repository when already loaded, or transparently loads it otherwise.

```{rave load_subject, language = "R", export = "subject"}
subject <- ravecore::new_rave_subject(
  project_name = project_name,
  subject_code = subject_code,
  strict = FALSE
)
```

```{rave clean_loaded_electrodes, language = "R", export = "loaded_electrode_clean"}
loaded_electrode_clean <- dipsaus::parse_svec(loaded_electrodes)
if (!length(loaded_electrode_clean)) {
  # blank -> all LFP channels
  loaded_electrode_clean <- subject$electrodes[subject$electrode_types == "LFP"]
}
loaded_electrode_clean <- sort(unique(as.integer(
  loaded_electrode_clean[loaded_electrode_clean %in% subject$electrodes]
)))
if (!length(loaded_electrode_clean)) {
  stop("No valid electrodes to load for this subject.")
}
```

```{rave clean_analysis_electrodes, language = "R", export = "analysis_electrode_clean"}
stim_e <- dipsaus::parse_svec(stim_electrode)
requested <- dipsaus::parse_svec(analysis_electrodes)
if (!length(requested)) {
  # blank -> all loaded LFP contacts except the stim electrode(s)
  requested <- setdiff(loaded_electrode_clean, stim_e)
}
lfp <- subject$electrodes[subject$electrode_types == "LFP"]
analysis_electrode_clean <- sort(unique(as.integer(
  requested[requested %in% loaded_electrode_clean & requested %in% lfp]
)))
```

```{rave clean_blocks, language = "R", export = "loaded_block_clean"}
loaded_block_clean <- loaded_blocks
if (!length(loaded_block_clean) ||
    (length(loaded_block_clean) == 1 && is.na(loaded_block_clean[[1]]))) {
  loaded_block_clean <- subject$blocks
}
loaded_block_clean <- intersect(as.character(unlist(loaded_block_clean)), subject$blocks)
if (!length(loaded_block_clean)) {
  loaded_block_clean <- subject$blocks
}
if (!isTRUE(as.character(analysis_block) %in% loaded_block_clean)) {
  stop(sprintf("analysis_block '%s' is not among the loaded blocks: %s",
               analysis_block, paste(loaded_block_clean, collapse = ", ")))
}
```

```{rave load_repository, language = "R", export = "repository"}
repository <- ravecore::prepare_subject_raw_voltage_with_blocks(
  subject = subject,
  electrodes = loaded_electrode_clean,
  blocks = loaded_block_clean,
  downsample = NA
)
```

```{rave load_signals, language = "R", export = "loaded_signals"}
# Materialize each loaded block as a (Time x Electrode) filearray handle.
filearray_root <- file.path(pipeline$pipeline_path, "data", "raw-voltage")
container <- repository$get_container()

loaded_signals <- structure(
  names = loaded_block_clean,
  lapply(loaded_block_clean, function(block) {
    ex <- extract_block_signal(container[[block]], loaded_electrode_clean)
    new_rave_filearray(
      filebase = file.path(filearray_root, block),
      data = ex$data,
      dimnames_list = list(Time = ex$time, Electrode = loaded_electrode_clean),
      headers = list(
        subject_id = repository$subject$subject_id,
        block = block,
        sample_rate = ex$sample_rate,
        max_duration = max(ex$time)
      )
    )
  })
)
```

The stim electrode carries near-square pulses; here is its raw trace for the block being
analyzed. This is the signal we threshold to recover pulse timing (MATLAB `plot(sync)`).

```{r diag_stim_trace, echo=FALSE, fig.height=3.5}
try({
  stim <- load_stim_channel(repository, stim_electrode, block = analysis_block)
  plot_stim_trace_with_onsets(
    trace = stim$data, sample_rate = stim$sample_rate,
    threshold = as.numeric(stim_threshold),
    time_range = c(0, min(5, max(stim$time))),
    main = sprintf("Stim electrode %s - block %s (first 5 s)",
                   stim_electrode, analysis_block)
  )
})
```

# Section 2 — Pulse detection

We recover one onset per pulse, from two sources in priority order.

**Preferred — the Events / sync channel.** One of the subject's Auxiliary channels carries a
clean square wave that is **high for the entire stim train**; its rising/falling edges mark
train onset and offset (exactly as the MATLAB prototype used the "Events" channel). RAVE does
not preserve the original channel name (all labels become `NoLabel`), so we identify it by
**shape**: among the aux channels, the one with almost no samples at intermediate levels and a
plausible number of train edges (`auto_detect_event_channel`; overridable via `event_channel`).
Each train is then split into `stim_frequency × stim_train_duration` evenly spaced onsets with
`linspace` — the MATLAB method — after an auto (bimodal) threshold.

**Fallback — the stim electrode.** When no Events channel exists we threshold the stim
electrode's own saturating signal. It rails in both directions and each biphasic pulse crosses
threshold several times, so we threshold the **absolute value** (`stim_threshold`) and coalesce
crossings with a **refractory period** (0.8 × inter-pulse interval) to keep one onset per pulse.

Either way, onsets are detected once per block, shared across all channels. The per-pulse
window is `snip_window = round(sample_rate / stim_frequency)` (one inter-pulse interval).

```{rave detect_pulses, language = "R", export = "pulse_info"}
# Resolve the Events channel: explicit setting, else auto-detect the square-wave
# aux channel. Prefer it; fall back to thresholding the stim electrode.
ec <- suppressWarnings(as.integer(dipsaus::parse_svec(event_channel)))
ec <- ec[is.finite(ec)]
if (!length(ec)) {
  ec <- tryCatch(auto_detect_event_channel(subject, analysis_block),
                 error = function(e) NULL)
}

if (length(ec) && !is.na(ec[[1]]) && ec[[1]] %in% subject$electrodes) {
  ev <- load_stim_channel(repository, ec[[1]], block = analysis_block)
  info <- detect_pulses_from_events(
    events_signal = ev$data, sample_rate = ev$sample_rate,
    stim_frequency = as.numeric(stim_frequency),
    stim_train_duration = as.numeric(stim_train_duration),
    pulse_duration_sec = as.numeric(pulse_duration) / 1000
  )
  info$sample_rate <- ev$sample_rate
  info$detection_source <- "events"
  info$event_channel <- ec[[1]]
} else {
  stim <- load_stim_channel(repository, stim_electrode, block = analysis_block)
  info <- detect_pulses_from_stim_channel(
    signal = stim$data, sample_rate = stim$sample_rate,
    threshold = as.numeric(stim_threshold),
    stim_frequency = as.numeric(stim_frequency),
    refractory_frac = 0.8,
    pulse_duration_sec = as.numeric(pulse_duration) / 1000
  )
  info$sample_rate <- stim$sample_rate
  info$detection_source <- "stim_electrode"
  info$event_channel <- NA_integer_
}
info$block <- as.character(analysis_block)
info$stim_electrode <- stim_electrode
pulse_info <- info
```

The detected onsets (red) on the detection channel, zoomed to the first train, and the raw
per-pulse snippets cut at those onsets on the stim electrode (MATLAB `plot(snips)`) — before
alignment they are jittered by up to a few samples.

```{r diag_detection, echo=FALSE, fig.height=3.5}
try({
  if (isTRUE(pulse_info$n_pulses > 0)) {
    src <- pulse_info$detection_source
    det_ch <- if (identical(src, "events")) pulse_info$event_channel else stim_electrode
    chan <- load_stim_channel(repository, det_ch, block = analysis_block)
    sr <- chan$sample_rate
    first_on <- pulse_info$onset_index[[1]]
    plot_stim_trace_with_onsets(
      trace = chan$data, sample_rate = sr, onsets = pulse_info$onset_index,
      threshold = NULL,
      time_range = c(max(0, (first_on / sr) - 0.05), (first_on / sr) + 0.6),
      main = sprintf("%s %s + detected onsets (%d pulses · source: %s)",
                     if (identical(src, "events")) "Events channel" else "Stim electrode",
                     det_ch, pulse_info$n_pulses, src)
    )
  } else {
    plot.new(); text(0.5, 0.5, "No pulses detected (check event_channel / stim_threshold)")
  }
})
```

```{r diag_raw_snippets, echo=FALSE, fig.height=3.5}
try({
  stim <- load_stim_channel(repository, stim_electrode, block = analysis_block)
  sr <- stim$sample_rate
  if (isTRUE(pulse_info$n_pulses > 0)) {
    win <- suppressWarnings(as.integer(snip_window))
    if (!length(win) || is.na(win) || win < 2L) win <- round(sr / as.numeric(stim_frequency))
    raw_snips <- extract_snippets(stim$data, pulse_info$onset_index, snip_window = win, pre = 0)
    plot_pulse_snippets_overlay(
      raw_snips, sample_rate = sr, pre = 0,
      main = sprintf("Raw pulse snippets on stim electrode (n = %d, pre-alignment)",
                     ncol(raw_snips))
    )
  }
})
```

# Section 3 — Bipolar re-reference

A bipolar derivation subtracts two nearby contacts (`anode - cathode`). Because the
secondary artifact and common-mode noise are shared by neighbouring contacts, the
subtraction cancels most of it - while a genuine local brain signal, which differs between
contacts, survives. Two rules keep the derivation physically meaningful:

* **Stay within a shaft.** Contacts on different probes are not physically adjacent, so we
  never pair across shafts. The shaft is the electrode's surgical label with trailing
  digits removed (`LabelPrefix`, else `gsub("[0-9]+$","",Label)`).
* **Skip the stim contact.** The stim electrode is saturated and useless as a reference, so
  we remove it and pair its neighbours across the gap (e.g. stim 38 => bipolar **37-39**).

Channels far from the stim site are usually clean after this step. Near-site channels often
retain a residual secondary artifact, visible as sharp peaks at the stim frequency and its
harmonics in the Welch periodogram - those are the channels that need Section 4.

```{rave build_bipolar_pairs, language = "R", export = "bipolar_pairs"}
electrode_table <- subject$get_electrode_table()
stim_e <- dipsaus::parse_svec(stim_electrode)

bipolar_pairs <- build_bipolar_pairs(
  electrode_table = electrode_table,
  analysis_electrodes = analysis_electrode_clean,
  stim_electrodes = stim_e,
  available_electrodes = loaded_electrode_clean
)
```

```{rave apply_bipolar, language = "R", export = "bipolar_signals"}
block <- as.character(analysis_block)
arr <- loaded_signals[[block]]$`@impl`
sample_rate <- arr$get_header("sample_rate")
signals <- arr[, , drop = FALSE, dimnames = FALSE]      # Time x Electrode
time <- as.numeric(dimnames(arr)$Time)

bip <- apply_bipolar_pairs(signals, loaded_electrode_clean, bipolar_pairs)

filearray_root <- file.path(pipeline$pipeline_path, "data", "bipolar-voltage")
if (ncol(bip) > 0) {
  bipolar_signals <- new_rave_filearray(
    filebase = file.path(filearray_root, block),
    data = bip,
    dimnames_list = list(Time = time, Electrode = colnames(bip)),
    headers = list(block = block, sample_rate = sample_rate)
  )
} else {
  bipolar_signals <- NULL
}
```

Monopolar vs bipolar Welch periodogram for a near-stim channel. The stim-frequency comb
(dashed grey) should shrink after bipolar re-referencing.

```{r diag_pwelch_bipolar, echo=FALSE, fig.height=4}
try({
  block <- as.character(analysis_block)
  if (!is.null(bipolar_signals) && nrow(bipolar_pairs) > 0) {
    barr <- bipolar_signals$`@impl`
    sr <- barr$get_header("sample_rate")
    labs <- dimnames(barr)$Electrode
    # prefer a bridged (across-stim) pair - the closest to the stim site
    idx <- which(bipolar_pairs$bridged)
    disp <- if (length(idx)) bipolar_pairs$label[[idx[[1]]]] else labs[[1]]
    disp_i <- match(disp, labs)
    if (is.na(disp_i)) disp_i <- 1L
    bipolar_trace <- barr[, disp_i, drop = TRUE, dimnames = FALSE]

    # monopolar anode of the same pair
    anode <- bipolar_pairs$anode[match(labs[[disp_i]], bipolar_pairs$label)]
    marr <- loaded_signals[[block]]$`@impl`
    mono_i <- match(anode, loaded_electrode_clean)
    mono_trace <- marr[, mono_i, drop = TRUE, dimnames = FALSE]

    harmonics <- as.numeric(stim_frequency) * seq_len(10)
    plot_pwelch_compare(
      traces = stats::setNames(
        list(mono_trace, bipolar_trace),
        c(sprintf("monopolar %s", anode), sprintf("bipolar %s", labs[[disp_i]]))),
      sample_rate = sr, win_sec = 1, max_freq = 500, marks = harmonics,
      main = "Monopolar vs bipolar periodogram (near-stim channel)"
    )
  } else {
    plot.new(); text(0.5, 0.5, "No within-shaft bipolar pairs for this selection")
  }
})
```

# Section 4 — Template subtraction (near-stim channels)

The two artifacts are removed by two tiers of cleaning. **Every** bipolar channel has its
saturated **direct** artifact blanked and interpolated (it rails on all channels and carries
no usable signal). The **secondary** rebound is handled by tier: channels **far** from the
stim site are already clean after bipolar re-referencing and need nothing more, while channels
**near** the site need per-pulse **template subtraction**. You pick the near channels via
`template_electrodes` (a pair qualifies when either contact is listed) — typically guided by
the Section 3 periodogram; `~` treats all pairs as near.

For near channels the template is estimated and subtracted (MATLAB L73-199):

1. **Snip & align.** Cut one inter-pulse window per pulse at the detected onsets, then
   align them by their dominant local extremum in rows `align_search_start:align_search_end`
   (the polarity - trough or crest - is chosen by whichever has the larger mean magnitude),
   shifting each onset by `ix - ix[1]` so the pulses superimpose.
2. **Blank.** Set the first `artifact_blank_width` samples (the saturated direct artifact)
   and the final seam sample to `NA`; these are excluded from the template and interpolated
   later.
3. **Profile / cluster.** Run PCA on the pulses and (optionally) cluster them with
   `profile_pulses(k = n_clusters)`. With `n_clusters = 1` this is the single global
   template of the MATLAB script; `k > 1` captures distinct artifact shapes (e.g. drift or
   state changes) with one template each. The function is method-pluggable so alternative
   clusterers can be dropped in later.
4. **Template.** Each cluster's template is the demeaned (`detrend(..., 0)`) average of its
   pulses over the non-blanked rows.
5. **Subtract, back-fill, gap-fill.** Subtract each pulse's cluster template on the valid
   rows, write the corrected snippets back into the continuous trace (leaving the blanked
   region as `NA`), then fill those gaps with a centered moving average
   (`gap_fill_window`) followed by nearest-value for any edge residue.

Far channels skip steps 3-5: they get only the direct-artifact blank + interpolation
(`blank_direct_artifact_and_fill`). The per-channel cleaning method is recorded on each
cleaned trace's header (`bipolar+template` vs `bipolar+direct`).

```{rave build_snippets, language = "R", export = "snippet_list"}
block <- as.character(analysis_block)
sample_rate <- pulse_info$sample_rate
pre <- abs(as.integer(template_window_pre))
win <- suppressWarnings(as.integer(snip_window))
if (!length(win) || is.na(win) || win < 2L) win <- round(sample_rate / as.numeric(stim_frequency))

snippet_list <- list()
if (!is.null(bipolar_signals) && isTRUE(pulse_info$n_pulses > 0)) {
  barr <- bipolar_signals$`@impl`
  labs <- dimnames(barr)$Electrode
  snippet_list <- structure(
    names = labs,
    lapply(seq_along(labs), function(j) {
      trace <- barr[, j, drop = TRUE, dimnames = FALSE]
      list(
        snips = extract_snippets(trace, pulse_info$onset_index, snip_window = win, pre = pre),
        onset_index = pulse_info$onset_index,
        pre = pre, snip_window = win, sample_rate = sample_rate
      )
    })
  )
}
```

```{rave align_pulses, language = "R", export = "aligned_pulse_list"}
search_start <- as.integer(align_search_start)
search_end <- as.integer(align_search_end)
pulse_dur_samples <- round((as.numeric(pulse_duration) / 1000) * pulse_info$sample_rate)

aligned_pulse_list <- list()
if (length(snippet_list) && !is.null(bipolar_signals)) {
  barr <- bipolar_signals$`@impl`
  labs <- names(snippet_list)
  aligned_pulse_list <- structure(
    names = labs,
    lapply(seq_along(labs), function(j) {
      snip <- snippet_list[[j]]
      trace <- barr[, j, drop = TRUE, dimnames = FALSE]

      al <- align_pulses_minmax(
        signal = trace, onset_index = snip$onset_index, snip_window = snip$snip_window,
        search_start = search_start, search_end = search_end,
        pulse_duration_samples = pulse_dur_samples
      )
      snips <- extract_snippets(trace, al$onset_index, snip_window = snip$snip_window, pre = snip$pre)
      list(
        snips = snips, onset_index = al$onset_index, n_pulses = al$n_pulses,
        pre = snip$pre, snip_window = snip$snip_window, sample_rate = snip$sample_rate
      )
    })
  )
}
```

```{rave compute_template, language = "R", export = "template_list"}
k <- max(1L, as.integer(n_clusters))
blank_width <- as.integer(artifact_blank_width)
ncomp <- max(1L, as.integer(pca_n_components))

template_list <- list()
if (length(aligned_pulse_list)) {
  labs <- names(aligned_pulse_list)
  template_list <- structure(
    names = labs,
    lapply(seq_along(labs), function(j) {
      al <- aligned_pulse_list[[j]]
      if (!isTRUE(ncol(al$snips) > 0)) return(NULL)

      n_rows <- nrow(al$snips)
      valid_rows <- setdiff(seq_len(n_rows), unique(c(seq_len(blank_width), n_rows)))

      profile <- profile_pulses(
        al$snips, k = k, method = "pca_kmeans",
        n_components = ncomp, valid_rows = valid_rows
      )
      ctempl <- compute_cluster_templates(al$snips, profile$cluster, blank_width = blank_width)
      template_full <- vapply(ctempl$templates, function(t) t$template_full, numeric(n_rows))

      list(
        profile = profile,
        cluster_templates = ctempl,
        template_full = matrix(template_full, nrow = n_rows,
                               dimnames = list(NULL, names(ctempl$templates))),
        blank_rows = ctempl$blank_rows,
        sample_rate = al$sample_rate, pre = al$pre
      )
    })
  )
}
```

```{rave clean_signals, language = "R", export = "cleaned_signals"}
block <- as.character(analysis_block)
gap_win <- as.integer(gap_fill_window)
blank_width <- as.integer(artifact_blank_width)
filearray_root <- file.path(pipeline$pipeline_path, "data", "cleaned-voltage")

# Two-tier: EVERY channel gets direct-artifact blank + interpolate; template
# subtraction runs only on the near-stim channels. A pair is "near" when either
# contact is in `template_electrodes` (blank => all pairs are near).
te <- dipsaus::parse_svec(template_electrodes)
pair_is_near <- function(lab) {
  if (!length(te)) { return(TRUE) }
  row <- bipolar_pairs[bipolar_pairs$label == lab, , drop = FALSE]
  if (!nrow(row)) { return(TRUE) }
  isTRUE(row$anode[[1]] %in% te) || isTRUE(row$cathode[[1]] %in% te)
}

cleaned_signals <- list()
if (length(aligned_pulse_list) && !is.null(bipolar_signals)) {
  barr <- bipolar_signals$`@impl`
  labs <- names(aligned_pulse_list)
  time <- as.numeric(dimnames(barr)$Time)

  cleaned_signals <- structure(
    names = labs,
    lapply(seq_along(labs), function(j) {
      lab <- labs[[j]]
      al <- aligned_pulse_list[[j]]
      tmpl <- template_list[[lab]]
      trace <- barr[, j, drop = TRUE, dimnames = FALSE]

      near <- pair_is_near(lab)
      if (!isTRUE(al$n_pulses > 0)) {
        cleaned <- as.numeric(trace)
        method <- "none"
      } else if (near && !is.null(tmpl)) {
        cleaned <- subtract_cluster_templates_and_fill(
          signal = trace, snips = al$snips, onset_index = al$onset_index,
          pre = al$pre, cluster_templates = tmpl$cluster_templates,
          gap_fill_window = gap_win
        )
        method <- "bipolar+template"
      } else {
        cleaned <- blank_direct_artifact_and_fill(
          signal = trace, onset_index = al$onset_index, snip_window = al$snip_window,
          blank_width = blank_width, pre = al$pre, gap_fill_window = gap_win
        )
        method <- "bipolar+direct"
      }

      new_rave_filearray(
        filebase = file.path(filearray_root, sprintf("%s_%s", block, lab)),
        data = matrix(cleaned, ncol = 1L),
        dimnames_list = list(Time = time, Electrode = lab),
        headers = list(
          block = block, channel = lab,
          sample_rate = barr$get_header("sample_rate"),
          n_pulses = al$n_pulses, cleaning_method = method
        )
      )
    })
  )
}
```

## Diagnostic figures

The following figures are drawn for the near-stim display channel (the bridged pair when
available). They reproduce, in order, the key MATLAB visualizations.

```{r diag_display_channel, echo=FALSE}
.disp <- NULL
try({
  if (length(template_list)) {
    labs <- names(template_list)
    idx <- which(bipolar_pairs$bridged & bipolar_pairs$label %in% labs)
    .disp <- if (length(idx)) bipolar_pairs$label[[idx[[1]]]] else labs[[1]]
  }
})
```

**Aligned pulses with the estimated template(s).** Grey/colored = individual aligned
pulses; heavy lines = the per-cluster templates (MATLAB `plot(snips2)` + template).

```{r diag_aligned_snippets, echo=FALSE, fig.height=3.8}
try({
  if (!is.null(.disp)) {
    al <- aligned_pulse_list[[.disp]]; tmpl <- template_list[[.disp]]
    plot_pulse_snippets_overlay(
      al$snips, sample_rate = al$sample_rate, pre = al$pre,
      template_full = tmpl$template_full, cluster = tmpl$profile$cluster,
      main = sprintf("Aligned pulses - channel %s (n = %d)", .disp, ncol(al$snips))
    )
  }
})
```

**Pulse PCA scores, colored by cluster** (MATLAB `scatter(Y2)`). Clear separation suggests
more than one artifact shape and motivates `n_clusters > 1`.

```{r diag_pca, echo=FALSE, fig.height=3.8}
try({
  if (!is.null(.disp)) {
    tmpl <- template_list[[.disp]]
    plot_pca_scatter(tmpl$profile, cluster = tmpl$profile$cluster)
  }
})
```

**Estimated template(s)** with the blanked direct-artifact region shaded (MATLAB
`plot(temp1)`).

```{r diag_template, echo=FALSE, fig.height=3.5}
try({
  if (!is.null(.disp)) {
    tmpl <- template_list[[.disp]]
    plot_template_waveform(
      tmpl$template_full, sample_rate = tmpl$sample_rate,
      blank_width = length(tmpl$blank_rows), pre = tmpl$pre
    )
  }
})
```

**Template-subtracted pulses** - the residual after removing each pulse's cluster template,
with the blanked artifact region left as `NA` (MATLAB `plot(psnips_sub)`).

```{r diag_subtracted, echo=FALSE, fig.height=3.5}
try({
  if (!is.null(.disp)) {
    al <- aligned_pulse_list[[.disp]]; tmpl <- template_list[[.disp]]
    ct <- tmpl$cluster_templates
    sub <- al$snips
    for (i in seq_len(ncol(sub))) {
      t_i <- ct$templates[[paste0("cluster", ct$cluster[[i]])]]
      sub[t_i$valid_rows, i] <- sub[t_i$valid_rows, i] - t_i$template_valid
      sub[t_i$blank_rows, i] <- NA_real_
    }
    plot_pulse_snippets_overlay(
      sub, sample_rate = al$sample_rate, pre = al$pre,
      template_full = rowMeans(sub, na.rm = TRUE),
      main = sprintf("Template-subtracted pulses - channel %s", .disp)
    )
  }
})
```

**Before vs after** on the continuous bipolar trace (MATLAB "Original vs cleaned").

```{r diag_before_after, echo=FALSE, fig.height=3.8}
try({
  if (!is.null(.disp)) {
    barr <- bipolar_signals$`@impl`
    labs <- dimnames(barr)$Electrode
    raw <- barr[, match(.disp, labs), drop = TRUE, dimnames = FALSE]
    cln <- cleaned_signals[[.disp]]$`@impl`[drop = TRUE, dimnames = FALSE]
    sr <- cleaned_signals[[.disp]]$`@impl`$get_header("sample_rate")
    plot_before_after(raw, cln, sample_rate = sr,
                      time_range = c(0, min(5, (length(raw) - 1) / sr)),
                      main = sprintf("Original vs cleaned - channel %s (first 5 s)", .disp))
  }
})
```

**Welch periodogram before vs after cleaning** - the stim-frequency comb should be further
attenuated relative to the bipolar-only trace.

```{r diag_pwelch_clean, echo=FALSE, fig.height=4}
try({
  if (!is.null(.disp)) {
    barr <- bipolar_signals$`@impl`
    labs <- dimnames(barr)$Electrode
    raw <- barr[, match(.disp, labs), drop = TRUE, dimnames = FALSE]
    cln <- cleaned_signals[[.disp]]$`@impl`[drop = TRUE, dimnames = FALSE]
    sr <- cleaned_signals[[.disp]]$`@impl`$get_header("sample_rate")
    harmonics <- as.numeric(stim_frequency) * seq_len(10)
    plot_pwelch_compare(
      traces = stats::setNames(list(raw, cln), c("bipolar", "cleaned")),
      sample_rate = sr, win_sec = 1, max_freq = 500, marks = harmonics,
      main = sprintf("Bipolar vs cleaned periodogram - channel %s", .disp)
    )
  }
})
```

# Export — write the cleaned data back as a reference

The point of this pipeline is that the cleaned recording can be analyzed **exactly like a
no-stim recording**. RAVE achieves that through the reference system: a downstream module
loads a channel as `raw(electrode) − reference`, so we store, for each bipolar pair's anode,
the reference `raw(anode) − cleaned`. Then selecting this reference at load time yields the
cleaned bipolar signal — no stim-specific handling anywhere downstream.

`export_cleaned_reference()` writes the reference signals to
`reference_path/ref_<name>_<a>_<c>/<block>/voltage` (the same on-disk format as
`ravecore::generate_reference`) and the reference table to `meta/reference_<name>.csv`. This
**writes into the subject's data directory**, so it is a deliberate, user-triggered step (the
module's export button), *not* something the knit does automatically — hence the chunk below is
`eval = FALSE`. Run it (or press Export in the app) once you are happy with the cleaning.

```{r export, eval=FALSE}
export_cleaned_reference(
  subject         = subject,
  cleaned_signals = cleaned_signals,
  bipolar_pairs   = bipolar_pairs,
  loaded_signals  = loaded_signals,
  block           = analysis_block,
  reference_name  = export_name
)
# Downstream then loads the cleaned data as an ordinary re-referenced subject, e.g.
#   ravecore::prepare_subject_raw_voltage_with_blocks(
#     subject, electrodes = <anodes>, reference_name = export_name)
```

# Build, Visualize, & Run

Please make sure the following code block is at the end of your pipeline file. This block
builds the pipeline and generates a `make-stim_artifact_removal.R` script.

```{r build, echo=FALSE, results='hide'}
build_pipeline(make_file = "make-stim_artifact_removal.R")
```

```{r visualize, echo=FALSE}
# Fixed usage, show pipeline graph
try({
  ravepipeline:::pipeline_dependency_graph(
    pipeline_path = ".", glimpse = TRUE)
}, silent = TRUE)
```
