---
title: "Remove Stimulation Artifacts"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
# This code block sets up the engine environment
# Please do not remove me
ravepipeline::pipeline_setup_rmd("stim_artifact_removal")
```

## From the loading screen

```{rave load_subject, export = "subject"}
subject <- ravecore::new_rave_subject(
  project_name = project_name,
  subject_code = subject_code
)
```


## Workflow




## Obsolete AI-generated non-sense

The pipeline ports the MATLAB stimulation-artifact-removal prototype
(`tmp/artifact_rej.m`) into RAVE targets. Detection and alignment reuse
`ravetools::stimpulse_find` / `ravetools::stimpulse_align`; the distinctive
template-subtraction + moving-average gap-fill cleaning lives in
`R/shared-signal_processing.R`. Every compute target loops over
`subject$blocks` and returns a named list keyed by block.

```{rave clean_loaded_electrodes, export = "loaded_electrodes_cleaned"}
loaded_electrodes <- ravecore:::parse_svec(loaded_electrodes)
if (!length(loaded_electrodes) %in% c(1, 2)) {
  stop("Only two electrode channels are allowed. Please enter one channel, or two channels (for bipolar re-reference)")
}

electrode_types <- subject$electrode_types[subject$electrodes %in% loaded_electrodes]
electrode_types <- unique(electrode_types)
if (length(electrode_types) != 1) {
  stop("Electrode channels must have the same signal type (e.g. both are LFP, or Aux, but not either).")
}
loaded_electrodes_cleaned <- sort(loaded_electrodes)
```

```{rave load_repository, export = "repository"}
# One repository loads the recording channels AND (when used) the trigger/sync
# channel together. The trigger is only added when detection is not forced to
# 'epoch' and a valid, distinct trigger electrode is provided.
elecs <- loaded_electrodes_cleaned
te <- tryCatch(as.integer(ravecore:::parse_svec(trigger_electrode)),
               error = function(e) integer(0))
use_trigger <- !identical(detection_source, "epoch") &&
  length(te) == 1 && te %in% subject$electrodes &&
  !te %in% loaded_electrodes_cleaned
if (use_trigger) {
  elecs <- sort(unique(c(loaded_electrodes_cleaned, te)))
}

repository <- ravecore::prepare_subject_raw_voltage_with_blocks(
  subject = subject,
  electrodes = elecs,
  downsample = NA
)
```

```{rave preload_stim_epoch, export = "preloaded_stim_epoch"}
if (isTRUE(epoch_name %in% subject$epoch_names)) {
  preloaded_stim_epoch <- subject$get_epoch(epoch_name)
} else {
  preloaded_stim_epoch <- NULL
}
```

```{rave load_signals, export = "loaded_signals"}
filearray_root <- file.path(pipeline$pipeline_path, "data", "raw-voltage")
if (!dir.exists(filearray_root)) {
  dir.create(filearray_root, showWarnings = FALSE, recursive = TRUE)
}

container <- repository$get_container()
rec_elecs <- loaded_electrodes_cleaned

loaded_signals <- lapply(subject$blocks, function(block) {
  fbase <- file.path(filearray_root, block)

  extracted <- extract_block_signal(container[[block]], rec_elecs)
  signals <- extracted$data
  sample_rate <- extracted$sample_rate
  time <- extracted$time

  if (file.exists(fbase)) {
    unlink(fbase, recursive = TRUE)
  }
  farr <- filearray::filearray_create(
    filebase = fbase,
    dimension = c(nrow(signals), 1L),
    type = "float",
    partition_size = 1L,
    initialize = FALSE
  )

  farr$set_header(key = "subject_id", value = repository$subject$subject_id, save = FALSE)
  farr$set_header(key = "block", value = block, save = FALSE)
  farr$set_header(key = "sample_rate", value = sample_rate, save = FALSE)
  farr$set_header(key = "max_duration", value = max(time), save = FALSE)

  # set epoch
  if (!is.null(preloaded_stim_epoch)) {
    epoch_table <- preloaded_stim_epoch$table
    if (is.data.frame(epoch_table) && nrow(epoch_table)) {
      block_epoch_table <- epoch_table[as.character(epoch_table$Block) == as.character(block), ]
      if (nrow(block_epoch_table)) {
        farr$set_header("epoch_table", block_epoch_table, save = FALSE)
      }
    }
  }

  if (length(rec_elecs) == 1) {
    farr[] <- signals[, 1]
    dimnames(farr) <- list(
      Time = time,
      Electrode = rec_elecs
    )
  } else {
    # bipolar = ch1 - ch2 (MATLAB convention)
    farr[] <- signals[, 1] - signals[, 2]
    dimnames(farr) <- list(
      Time = time,
      Electrode = sprintf("%s-%s", rec_elecs[[1]], rec_elecs[[2]])
    )
  }

  ravepipeline::RAVEFileArray$new(x = farr, temporary = FALSE)
})
names(loaded_signals) <- subject$blocks
```

```{rave load_trigger_signals, export = "trigger_signals"}
# Read the trigger/sync channel from the SAME repository container (no second
# repository). Populated only when a valid, distinct trigger electrode is used.
trigger_signals <- NULL
te <- tryCatch(as.integer(ravecore:::parse_svec(trigger_electrode)),
               error = function(e) integer(0))
use_trigger <- !identical(detection_source, "epoch") &&
  length(te) == 1 && te %in% repository$electrode_list &&
  !te %in% loaded_electrodes_cleaned

if (use_trigger) {
  trigger_root <- file.path(pipeline$pipeline_path, "data", "trigger-voltage")
  if (!dir.exists(trigger_root)) {
    dir.create(trigger_root, showWarnings = FALSE, recursive = TRUE)
  }
  container <- repository$get_container()

  trigger_signals <- lapply(subject$blocks, function(block) {
    extracted <- extract_block_signal(container[[block]], te)
    sig <- as.numeric(extracted$data[, 1])

    fbase <- file.path(trigger_root, block)
    if (file.exists(fbase)) { unlink(fbase, recursive = TRUE) }
    farr <- filearray::filearray_create(
      filebase = fbase, dimension = c(length(sig), 1L),
      type = "float", partition_size = 1L, initialize = FALSE)
    farr[] <- sig
    farr$set_header("block", block, save = FALSE)
    farr$set_header("sample_rate", extracted$sample_rate, save = FALSE)
    ravepipeline::RAVEFileArray$new(x = farr, temporary = FALSE)
  })
  names(trigger_signals) <- subject$blocks
}
```

```{rave detect_pulses, export = "pulse_info_list"}
pulse_duration_sec <- pulse_duration / 1000

# Resolve detection mode. 'auto' uses the trigger channel when available,
# otherwise the preloaded stim epoch. Neither uses ravetools.
trigger_available <- is.list(trigger_signals) && length(trigger_signals) > 0
detection_mode <- if (identical(detection_source, "trigger")) {
  "trigger"
} else if (identical(detection_source, "epoch")) {
  "epoch"
} else if (trigger_available) {
  "trigger"
} else {
  "epoch"
}

pulse_info_list <- structure(
  names = subject$blocks,
  lapply(subject$blocks, function(block) {
    arr <- loaded_signals[[block]]$`@impl`
    sample_rate <- arr$get_header("sample_rate")

    if (identical(detection_mode, "trigger") && trigger_available) {
      trig <- trigger_signals[[block]]$`@impl`[drop = TRUE, dimnames = FALSE]
      detect_pulses_from_trigger(
        trigger = trig, sample_rate = sample_rate,
        threshold = as.numeric(trigger_threshold),
        stim_frequency = stim_frequency, stim_length = stim_length,
        pulse_duration_sec = pulse_duration_sec)
    } else {
      epoch_table <- if (!is.null(preloaded_stim_epoch)) preloaded_stim_epoch$table else NULL
      detect_pulses_from_epoch(
        epoch_table = epoch_table, block = block,
        sample_rate = sample_rate, pulse_duration_sec = pulse_duration_sec)
    }
  })
)
```

```{rave align_pulses, export = "aligned_pulse_list"}
# Faithful MATLAB alignment: local min/max within rows search_start:search_end.
search_start <- as.integer(align_search_start)
search_end <- as.integer(align_search_end)

aligned_pulse_list <- structure(
  names = subject$blocks,
  lapply(subject$blocks, function(block) {
    info <- pulse_info_list[[block]]
    if (!isTRUE(info$n_pulses > 0)) { return(info) }

    arr <- loaded_signals[[block]]$`@impl`
    sample_rate <- arr$get_header("sample_rate")
    signal <- arr[drop = TRUE, dimnames = FALSE]

    win <- suppressWarnings(as.integer(snip_window))
    if (!length(win) || !isTRUE(win >= 2)) {
      win <- round(sample_rate / stim_frequency)
    }
    pulse_dur_samples <- round((pulse_duration / 1000) * sample_rate)

    align_pulses_minmax(
      signal, onset_index = info$onset_index, snip_window = win,
      search_start = search_start, search_end = search_end,
      pulse_duration_samples = pulse_dur_samples)
  })
)
```

```{rave build_snippets, export = "snippet_list"}
snippet_list <- structure(
  names = subject$blocks,
  lapply(subject$blocks, function(block) {
    info <- aligned_pulse_list[[block]]
    if (!isTRUE(info$n_pulses > 0)) { return(NULL) }

    arr <- loaded_signals[[block]]$`@impl`
    sample_rate <- arr$get_header("sample_rate")
    signal <- arr[drop = TRUE, dimnames = FALSE]

    win <- suppressWarnings(as.integer(snip_window))
    if (!length(win) || !isTRUE(win >= 2)) {
      win <- round(sample_rate / stim_frequency)
    }
    pre <- abs(as.integer(template_window_pre))

    snips <- extract_snippets(
      signal, onset_index = info$onset_index,
      snip_window = win, pre = pre)

    list(
      snips = snips,
      onset_index = info$onset_index,
      pre = pre,
      snip_window = win,
      sample_rate = sample_rate
    )
  })
)
```

```{rave compute_template, export = "template_list"}
template_list <- structure(
  names = subject$blocks,
  lapply(subject$blocks, function(block) {
    snip <- snippet_list[[block]]
    if (is.null(snip)) { return(NULL) }

    tmpl <- compute_stim_template(snip$snips, blank_width = artifact_blank_width)

    pca <- NULL
    if (isTRUE(pca_enabled)) {
      pca <- snippet_pca(
        snip$snips, valid_rows = tmpl$valid_rows,
        n_components = pca_n_components)
    }

    list(
      template = tmpl,
      pca = pca,
      sample_rate = snip$sample_rate,
      pre = snip$pre
    )
  })
)
```

```{rave clean_signals, export = "cleaned_signals"}
filearray_root <- file.path(pipeline$pipeline_path, "data", "cleaned-voltage")
if (!dir.exists(filearray_root)) {
  dir.create(filearray_root, showWarnings = FALSE, recursive = TRUE)
}

cleaned_signals <- structure(
  names = subject$blocks,
  lapply(subject$blocks, function(block) {
    arr <- loaded_signals[[block]]$`@impl`
    sample_rate <- arr$get_header("sample_rate")
    signal <- arr[drop = TRUE, dimnames = FALSE]

    info <- aligned_pulse_list[[block]]
    snip <- snippet_list[[block]]
    tmpl <- template_list[[block]]

    cleaned <- as.numeric(signal)
    n_pulses <- 0L
    if (!is.null(snip) && !is.null(tmpl) && isTRUE(info$n_pulses > 0)) {
      n_pulses <- info$n_pulses
      cleaned <- subtract_template_and_fill(
        signal = signal,
        snips = snip$snips,
        onset_index = snip$onset_index,
        pre = snip$pre,
        template = tmpl$template,
        gap_fill_window = as.integer(gap_fill_window))
    }

    fbase <- file.path(filearray_root, block)
    if (file.exists(fbase)) { unlink(fbase, recursive = TRUE) }
    farr <- filearray::filearray_create(
      filebase = fbase,
      dimension = c(length(cleaned), 1L),
      type = "float",
      partition_size = 1L,
      initialize = FALSE)
    farr[] <- cleaned
    dimnames(farr) <- dimnames(arr)

    farr$set_header("subject_id", arr$get_header("subject_id"), save = FALSE)
    farr$set_header("block", block, save = FALSE)
    farr$set_header("sample_rate", sample_rate, save = FALSE)
    farr$set_header("max_duration", arr$get_header("max_duration"), save = FALSE)
    farr$set_header("n_pulses", n_pulses, save = FALSE)
    farr$set_header("cleaning_method", "template_subtraction", save = FALSE)

    ravepipeline::RAVEFileArray$new(x = farr, temporary = FALSE)
  })
)
```

The following plain `r` block reproduces the MATLAB "original vs cleaned"
comparison during interactive knitting. It is **not** a pipeline target.

```{r before_after, echo=FALSE}
block <- subject$blocks[[1]]
raw <- loaded_signals[[block]]$`@impl`[drop = TRUE, dimnames = FALSE]
cln <- cleaned_signals[[block]]$`@impl`[drop = TRUE, dimnames = FALSE]
sr <- cleaned_signals[[block]]$`@impl`$get_header("sample_rate")
plot_before_after(raw, cln, sample_rate = sr, time_range = c(NA, NA))
```

## Build, Visualize, & Run

Please make sure the following code block is at the end of your pipeline file.
This block will build the pipeline and generate a
`make-stim_artifact_removal.R` script.

```{r build, echo=FALSE, results='hide'}
build_pipeline(make_file = "make-stim_artifact_removal.R")
```

```{r visualize, echo=FALSE}
# Fixed usage, show pipeline graph
try({
  ravepipeline:::pipeline_dependency_graph(
    pipeline_path = ".", glimpse = TRUE)
}, silent = TRUE)
```
